A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-09-26 doc/paper1-submission/paper2.tex

\documentclass[fleqn,12pt,english]{wlpeerj}
\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
%\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
%\usepackage{xcolor}
%\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}

@@

%\def\fps@figure{ht}

@@

\usepackage{babel}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\author[1,2]{Benjamin D. Redelings}
\author[2,3,4]{Mark T. Holder}
\affil[1]{Department of Biology, Duke University, Durham NC, US}
\affil[2]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
\affil[3]{Biodiversity Institute, University of Kansas, Lawrence KS, US}
\affil[4]{Heidelberg Institute for Theoretical Studies, Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany}
\corrauthor[2,3,4]{Mark T. Holder}{mtholder@ku.edu}

% \keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
We present a new supertree method and software implementation that
enables rapid estimation of a summary tree on the scale of millions
of leaves. This software pipeline is called \emph{``propinquity.''}
It uses the dependency-tracking tool \emph{make} to direct the series
of operations required to construct the tree. It relies heavily on
\emph{``otcetera''} - a set of C++ tools to performs most of the
steps of the pipeline. All of the components are free software and
are available on GitHub. This pipeline is currently used by the Open
Tree of Life project to produce all of the versions of project's ``synthetic
tree'' starting at version 5. The propinquity-based pipeline improves
on previous supertree construction methods because it produces no
unsupported branches, and it avoids unnecessary polytomies. In addition
to producing a tree, the pipeline writes an annotations file that
reports which grouping in the input trees support each of the groups
in the supertree.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

@@

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__abc}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__acd}

}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__bcd}

@@

\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s12}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s13}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s23}

@@

\caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
A trivial statemen from the first tree that a taxon labelled $ABCD$
is sister to $E$ has been omitted, because trees with only 2 leaves
do not contain phylogenetic information.}
\label{fig:subproblems-example}
\end{figure}
\begin{figure}
\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__ab1c}

}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__acb2}

}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__tax}

@@

\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__opt}

}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__returned}

@@

\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Figures__subproblem_solutions__ott14\lyxdot tre}

}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.55]{Figures__subproblem_solutions__ott17\lyxdot tre}

@@

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.38]{example3__cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.38]{example3__grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.38]{example3__unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.38]{example3__unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R and instead attached at the MRCA of R1, R2, and
R3.}
\end{figure}

@@

\hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
\shadowbox{\begin{minipage}[t]{1.0\columnwidth}%
\texttt{\{}

@@

@@

Here we have described the motivation and methodology used by the
propinquity tool that is currently used by the Open Tree of Life project
to build its summary tree. A modified version of the treemachine software
which built the summary tree described in the \citet{HinchliffEtAl2015}
paper is used by the project to serve the tree produced by propinquity
via Open Tree of Life APIs. The migration of summary tree construction
from treemachine (used for version 4) to propinquity (for all versions
from v5.0 to present) has decreased the computational time required
to construct a supertree from several hours to about 8 minutes (after
some preprocessing steps which only have to be performed when the
input taxonomy changes). The amount of RAM required during tree construction
has also decreased substantially. We ran propinquity on the same input
trees used to construct version Open Tree version 4. Unlike the version
4 tree, the tree from propinquity contained no unsupported nodes or
unnecessary polytomies. The number of displayed input splits increased
from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
of input splits that do not conflict with the summary tree, but are
not incorporated dropped from 2718 to 0. The number of input splits
that conflict with the phylogeny dropped from 2760 to 1357, a decrease
of 1357, or 49.2\%. %

@@

\section{The maximum sum displayed groups' weighted scores criterion}\label{sec:The-maximum-sum}

@@

\section{Description of the decomposition algorithm of otc-uncontested-decompose}\label{sec:Decompose-algorithm}

>>>>

%% LyX 2.2.0 created this file.  For more info, see http:____www.lyx.org__.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{paper}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{xcolor}
\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

@@

\usepackage{babel}
\begin{document}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\author{Benjamin D. Redelings$^{1,2}$, Mark T. Holder$^{1,3}$}
\maketitle
\begin{abstract}
We present a new supertree method and software implementation that
enables rapid estimation of a summary tree on the scale of millions
of leaves. This software pipeline is called \emph{``propinquity.''}
It uses the dependency-tracking tool \emph{make} to direct the series
of operations required to construct the tree. It relies heavily on
\emph{``otcetera''} - a set of C++ tools to performs most of the
steps of the pipeline. All of the components are free software and
are available on GitHub. This pipeline is currently used by the Open
Tree of Life project to produce all of the versions of project's ``synthetic
tree'' starting at version 5. The propinquity-based pipeline improves
on previous supertree construction methods because it produces no
unsupported branches, and it avoids unnecessary polytomies. In addition
to producing a tree, the pipeline writes an annotations file that
reports which grouping in the input trees support each of the groups
in the supertree.
\end{abstract}

@@

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__abc}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__acd}

}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__bcd}

@@

\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s12}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s13}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s23}

@@

\caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
A trivial statemen from the first tree that a taxon labelled $ABCD$
is sister to $E$ has been omitted, because trees with only 2 leaves
do not contain phylogenetic information.}
\label{fig:subproblems-example}
\end{figure}
\begin{figure}
\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__ab1c}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__acb2}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__tax}

@@

\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__opt}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__returned}

@@

\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott14\lyxdot tre}

}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott17\lyxdot tre}

@@

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3__cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3__grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.4]{example3__unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.4]{example3__unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R and instead attached at the MRCA of R1, R2, and
R3.}
\end{figure}

@@

\hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
\shadowbox{\begin{minipage}[t]{0.8\columnwidth}%
\texttt{\{}

@@

\paragraph{Annotations }

@@

Here we have described the motivation and methodology used by the
propinquity tool that is currently used by the Open Tree of Life project
to build its summary tree. A modified version of the treemachine software
which built the summary tree described in the \citet{HinchliffEtAl2015}
paper is used by the project to serve the tree produced by propinquity
via Open Tree of Life APIs. The migration of summary tree construction
from treemachine (used for version 4) to propinquity (for all versions
from v5.0 to present) has decreased the computational time required
to construct a supertree from several hours to about 8 minutes (after
some preprocessing steps which only have to be performed when the
input taxonomy changes). The amount of RAM required during tree construction
has also decreased substantially. We ran propinquity on the same input
trees used to construct version Open Tree version 4. Unlike the version
4 tree, the tree from propinquity contained no unsupported nodes or
unnecessary polytomies. The number of displayed input splits increased
from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
of input splits that do not conflict with the summary tree, but are
not incorporated dropped from 2718 to 0. The number of input splits
that conflict with the phylogeny dropped from 2760 to 1357, a decrease
of 1357, or 49.2\%. %
\begin{comment}
ARE THESE numbers for the same set of inputs? or v4 to v5 (the latter
had the barnacles tree collection, and perhaps a few other tree differences?
\end{comment}

@@

\section{\label{sec:The-maximum-sum}The maximum sum displayed groups' weighted
scores criterion}

@@

\section{\label{sec:Decompose-algorithm}Description of the decomposition
algorithm of otc-uncontested-decompose}
